---
title: "Lecture 03: Advanced Likelihood Methods and Inference"
subtitle: PUBH 8878, Statistical Genetics
format: 
    revealjs:
        theme: [default, "custom.scss"]
        slide-number: true
        chalkboard: true
        html-math-method: katex
execute:
  freeze: auto
  echo: true
  warning: true
engine: knitr
knitr:
  opts_chunk: 
    R.options:
      width: 80
draft: true
---

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(viridis)
library(htmltools)
```

## Review: Likelihood Function Basics

From Lecture 1, we learned:
- Likelihood function: $L(\boldsymbol{\theta} | \text{data}) = P(\text{data} | \boldsymbol{\theta})$
- Maximum Likelihood Estimation (MLE)
- Simple one-parameter examples

Today: **Multi-parameter likelihood surfaces and complex genetic models**

## Multi-Parameter Likelihood Surfaces

### Penetrance Model with Two Parameters

Consider a dominant disease model with:
- Allele frequency: $p$
- Penetrance for heterozygotes: $f_1$
- Fixed: $f_0 = 0$ (unaffected homozygotes), $f_2 = 1$ (affected homozygotes)

Population disease risk: $K = p^2 \cdot 1 + 2p(1-p) \cdot f_1 + (1-p)^2 \cdot 0$

## Likelihood Surface Visualization

```{r, echo = FALSE}
# Generate likelihood surface for penetrance model
p_vals <- seq(0.01, 0.5, length.out = 50)
f1_vals <- seq(0.1, 1.0, length.out = 50)

# Example data: 100 individuals, 15 affected, assuming f0=0, f2=1
n <- 100
k <- 15

likelihood_surface <- expand.grid(p = p_vals, f1 = f1_vals) %>%
  mutate(
    K = p^2 * 1 + 2 * p * (1-p) * f1 + (1-p)^2 * 0,
    log_likelihood = dbinom(k, n, K, log = TRUE)
  )

ggplot(likelihood_surface, aes(x = p, y = f1, fill = log_likelihood)) +
  geom_tile() +
  geom_contour(aes(z = log_likelihood), color = "white", alpha = 0.7) +
  scale_fill_viridis_c(name = "Log\nLikelihood") +
  labs(
    x = "Allele Frequency (p)",
    y = "Penetrance (f₁)",
    title = "Likelihood Surface: Dominant Model with Reduced Penetrance"
  ) +
  theme_minimal(base_size = 14)
```

## Interpreting Likelihood Surfaces

::: incremental
- **Peak**: Maximum likelihood estimates (MLE)
- **Contours**: Lines of equal likelihood
- **Confidence Regions**: Areas within likelihood ratio thresholds
- **Ridge vs Peak**: Identifiability issues
- **Flatness**: Parameter uncertainty
:::

## Likelihood Ratio Tests

### Testing Nested Hypotheses

**Full Model**: $H_1: \text{both } p \text{ and } f_1 \text{ free}$
**Reduced Model**: $H_0: f_1 = 1 \text{ (complete penetrance)}$

Test statistic: $\Lambda = 2[\ell(\hat{\boldsymbol{\theta}}_1) - \ell(\hat{\boldsymbol{\theta}}_0)]$

Under $H_0$: $\Lambda \sim \chi^2_{df}$ where $df = $ difference in parameters

## Profile Likelihood

When interested in one parameter but others are nuisance:

$$L_{\text{profile}}(p) = \max_{f_1} L(p, f_1)$$

```{r, echo=FALSE}
# Calculate profile likelihood for p
profile_p <- likelihood_surface %>%
  group_by(p) %>%
  summarise(
    max_log_likelihood = max(log_likelihood),
    best_f1 = f1[which.max(log_likelihood)]
  )

ggplot(profile_p, aes(x = p, y = max_log_likelihood)) +
  geom_line(linewidth = 1.2, color = "blue") +
  labs(
    x = "Allele Frequency (p)",
    y = "Profile Log-Likelihood",
    title = "Profile Likelihood for Allele Frequency"
  ) +
  theme_minimal(base_size = 14)
```

## Information and Standard Errors

### Fisher Information Matrix

For parameters $\boldsymbol{\theta} = (\theta_1, \theta_2)$:

$$I_{ij}(\boldsymbol{\theta}) = -E\left[\frac{\partial^2 \ell}{\partial \theta_i \partial \theta_j}\right]$$

**Asymptotic standard errors**: $SE(\hat{\theta}_i) \approx \sqrt{[I^{-1}]_{ii}}$

**Confidence intervals**: $\hat{\theta}_i \pm 1.96 \cdot SE(\hat{\theta}_i)$

## Practical Considerations

::: incremental
1. **Computational Methods**: Newton-Raphson, EM algorithm, numerical optimization
2. **Starting Values**: Multiple random starts to avoid local maxima  
3. **Convergence**: Check gradient, Hessian, parameter stability
4. **Model Comparison**: AIC, BIC for non-nested models
5. **Diagnostics**: Residuals, goodness-of-fit tests
:::

## Applications in Genetics

### Complex Segregation Analysis
- Mode of inheritance parameters
- Penetrance functions
- Gene frequency estimation

### Linkage Analysis  
- Recombination fraction estimation
- Multi-point linkage
- Disease gene localization

### Association Studies
- Allelic vs genotypic tests
- Hardy-Weinberg equilibrium
- Population stratification

## Coming Up

**Next Lecture**: Bayesian inference methods
- Prior specification
- Posterior computation  
- MCMC algorithms
- Comparison with likelihood approaches